---
title: "Correlates of driving under the influence of cannabis"
subtitle: "Latent Class Analysis"
author: "Anna Schranz"
date: "2024-05-25"
output: 
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn = -1)
```

# Setup
```{r essentials}
# clean workspace
rm(list=ls())
packages <- c("data.table", "readxl", "tidyverse", "ggplot2", "poLCA", "reshape2", "vcd", "tidyLPA", "export") 

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
# Load packages
invisible(lapply(packages, library, character.only = TRUE))

# current date:
DATE <- format(Sys.Date(), "%Y%m%d")

# themes and options
options(scipen = 999)

# output folders
folder_table <- file.path("..", "..", "tables", "LCA")
folder_plot <- file.path("..", "..", "figures", "LCA")

if (!file.exists(folder_table)) {
  dir.create(folder_table, recursive = TRUE)
}

if (!file.exists(folder_plot)) {
  dir.create(folder_plot, recursive = TRUE)
}
```

# Data Import
```{r data_import}
# define paths
path_DE_reg <- file.path("..", "Data", "regular_users_GER_cleaned_v03_20240731.rds") #at least monthly users in GER
path_PLZ <- file.path("..", "Data", "plz and urban.rds") #PLZ and urbanization level

# read data
data_raw <- data.table(readRDS(file = path_DE_reg))
data_PLZ <- data.table(readRDS(file = path_PLZ))
```


# Data Preparation

## Selection of variables relevant for LCA
```{r select_vars}
data_sel <- data_raw %>% 
  dplyr::select("i_NUMBER", "PLZ", "Bundesland", "agegroup", "age_in_years", "sex", "edu_group", "prescription", "CAST_score_full", "BSSS_total", "DISTRESS_total", "speeding_total", "DUIC.INTOX", "DUIA", "DUIC.PEERS", "LEGAL.01", "DUICDUIA.PERCRISK.1", "DUICDUIA.PERCRISK.2", "DUICDUIA.PERCRISK.3", "DUICDUIA.PERCRISK.4", "medic_withprescr_only", "DRIVERLICENSE")

data_PLZ <- data_PLZ %>% dplyr::rename(PLZ = plz)

data_sel <- data_sel %>% filter(DRIVERLICENSE == "ja") #only participants who have a driver's license

```

## Match postcodes to official municipality code and RegioStaRGem5 classification to define urbanicity (urban vs. rural dwellers)
```{r match_plz}

data_sel <- merge(data_sel, data_PLZ, by = "PLZ", all.x = TRUE)

data_sel <- data_sel %>% mutate(urb_lab_en = case_when(
  PLZ == "" & Bundesland %in% c("HB", "HH", "BE") ~ "metropolis",
  TRUE ~ urb_lab_en),
  # dichotomize urbanization level into rural and urban
  urb_lab_en_bin = factor(case_when(
    urb_lab_en %in% c("metropolis", "large city") ~ 2,
    urb_lab_en %in% c("medium-sized city", "urban space", "small town/village area") ~ 1,
    TRUE ~ NA_real_),
    levels = c(1, 2),
    labels = c("rural", "urban")))
```

### Distribution of urbanicity
```{r urban_dist}
prop.table(table(data_sel$urb_lab_en))
table(data_sel$urb_lab_en)

#how many NAs in urb_lab_en?
table(is.na(data_sel$urb_lab_en), data_sel$DUIC.INTOX) # 15 NAs (7 without DUIC, 8 with DUIC)
```

## Tranformation of variables into binary factors

For this analysis, variables should be binary factors with levels 1 and 2.

```{r data_prep}
data <- data_sel %>%
  mutate(
    CUD_CAST = factor(case_when(
      CAST_score_full >= 7 ~ 2,
      CAST_score_full < 7 ~ 1,
      TRUE ~ NA_real_),
      levels = c(1, 2),
      labels = c("low", "high")),
    medic_withprescr_only = factor(case_when(
      medic_withprescr_only == 1 ~ 2,
      medic_withprescr_only == 0 ~ 1,
      TRUE ~ NA_real_),
      levels = c(1, 2),
      labels = c("no", "yes")),
    mental_illness = factor(case_when(
      DISTRESS_total > 12 ~ 2,
      DISTRESS_total <= 12 ~ 1,
      TRUE ~ NA_real_),
      levels = c(1, 2),
      labels = c("no", "yes")),
    LEGALPROBL = factor(case_when(
      LEGAL.01 == "ja" ~ 2,
      LEGAL.01 == "nein" ~ 1,
      TRUE ~ NA_real_),
      levels = c(1, 2),
      labels = c("no", "yes")),
    DUIA12m = factor(case_when(
      DUIA %in% c("in den letzten 30T", "in den letzten 12M") ~ 2,
      DUIA %in% c("nie", "länger als 12M her") ~ 1,
      is.na(DUIA) ~ 1, #NA = does not drink alcohol -> cannot drive a car while intoxicated
      TRUE ~ NA_real_), 
      levels = c(1, 2),
      labels = c("no", "yes")),
    DUICPEERS12m = factor(case_when(
      DUIC.PEERS %in% c("in den letzten 30T", "in den letzten 12M") ~ 2,
      DUIC.PEERS %in% c("nie", "länger als 12M her") ~ 1,
      TRUE ~ NA_real_),
      levels = c(1, 2),
      labels = c("no", "yes")),
    DUICINTOX12m = factor(case_when(
      DUIC.INTOX %in% c("in den letzten 30T", "in den letzten 12M") ~ 2,
      DUIC.INTOX %in% c("nie", "länger als 12M her") ~ 1,
      TRUE ~ NA_real_),
      levels = c(1, 2),
      labels = c("no", "yes")),
    PERCRISK_DUICmod_bin = factor(case_when(
      DUICDUIA.PERCRISK.1 %in% c("ungefährlich", "nicht so gefährlich") ~ 2,
      DUICDUIA.PERCRISK.1 %in% c("gefährlich", "sehr gefährlich") ~ 1,
      TRUE ~ NA_real_),
      levels = c(1, 2),
      labels = c("high risk perc", "low risk perc")),
    PERCRISK_DUIAmod_bin = factor(case_when(
      DUICDUIA.PERCRISK.2 %in% c("ungefährlich", "nicht so gefährlich") ~ 2,
      DUICDUIA.PERCRISK.2 %in% c("gefährlich", "sehr gefährlich") ~ 1,
      TRUE ~ NA_real_),
      levels = c(1, 2),
      labels = c("high risk perc", "low risk perc")))
    
```

## Median split to dichotomize continuous variables

### Brief Sensation Seeking Scale (BSSS)
```{r dichotomize_bsss}
#subtract -1 from BSSS_total, so the scale starts at 0 (to be consistent with the other scales)
data <- data %>% mutate(BSSS_total = BSSS_total - 1)

summary(data$BSSS_total) #median = 2.125
sum(is.na(data$BSSS_total)) #0 NAs

data <- data %>% 
  mutate(BSSS_binary = factor(case_when(
    BSSS_total <= median(BSSS_total) ~ 1,
    BSSS_total > median(BSSS_total) ~ 2,
    TRUE ~ NA_real_),
    levels = c(1, 2),
    labels = c("low", "high")))
```

### Speeding
```{r dichotomize_speeding}
# subtract -1 from speeding_total, so the scale starts at 0 (to be consistent with the other scales)
data <- data %>% mutate(speeding_total = speeding_total - 1)

summary(data$speeding_total) #median = 1.5, 27 NAs
median_speeding_total <- median(data$speeding_total, na.rm = TRUE)

data <- data %>% 
  mutate(speeding_binary = factor(case_when(
    speeding_total <= median_speeding_total ~ 1,
    speeding_total > median_speeding_total ~ 2,
    TRUE ~ NA_real_),
    levels = c(1, 2),
    labels = c("low", "high")))
```


## Description of the sample

### Sociodemographic variables (covariates)
```{r descriptives_demo}

#randomly assign sex == "divers" to "male" or "female"
divers_indices <- which(data$sex == "divers")
n_divers <- length(divers_indices)
set.seed(42)
new_values <- sample(c("weiblich", "männlich"), n_divers, replace = TRUE)
data$sex[divers_indices] <- new_values
data$sex <- factor(data$sex, levels = c("männlich", "weiblich"))


prop.table(table(data$urb_lab_en_bin, useNA = "ifany"))
prop.table(table(data$sex, useNA = "ifany"))
prop.table(table(data$edu_group, useNA = "ifany"))
prop.table(table(data$agegroup, useNA = "ifany"))

summary(data$age_in_years)
sd(data$age_in_years)

cat("missings in all covariates")
sum(is.na(data$sex) | is.na(data$agegroup) | is.na(data$edu_group) | is.na(data$urb_lab_en_bin))

``` 

### Indicators
```{r descriptives}
ind_vars <- c("DUIA12m", "DUICINTOX12m", "CUD_CAST", "mental_illness", "medic_withprescr_only", "LEGALPROBL", "DUICPEERS12m", "BSSS_binary", "speeding_binary", "PERCRISK_DUICmod_bin", "PERCRISK_DUIAmod_bin")

 for (var in ind_vars) {
   cat("Variable: ", var, "\n")
   print(prop.table(table(data[[var]], useNA = "ifany")))
   cat("\n\n")
 }

#CAST
dist_CAST <- ggplot(data, aes(x = CAST_score_full, fill = CUD_CAST)) +
  geom_histogram(bins = 20) +
  labs(x = "Cannabis Abuse Screening Test - Score", y = "Count", fill = "CUD risk") +
  scale_x_continuous(breaks = seq(0, 24, by = 2, limits = c(0, 24))) +
  theme_minimal()

dist_CAST

#PERCRISK_DUICmod_bin
data <- data %>% mutate(DUICDUIA.PERCRISK.1 = factor(DUICDUIA.PERCRISK.1, levels = c("ungefährlich", "nicht so gefährlich", "gefährlich", "sehr gefährlich", "weiß nicht"), labels = c("not dangerous", "not so dangerous", "dangerous", "very dangerous", "don't know")))

dist_PercRiskDUIC <- ggplot(data, aes(x = DUICDUIA.PERCRISK.1, fill = PERCRISK_DUICmod_bin)) +
  geom_bar() +
  labs(x = "How dangerous do you think it is\nto take 1-2 puffs on a joint before driving?", y = "Count", fill = "Dichotomized\nRisk\nPerception\nDUIC") +
  scale_fill_discrete(labels = c("high risk perc" = "High", "low risk perc" = "Low")) +
  scale_x_discrete(labels = c("not dangerous" = "Not\ndangerous", "not so dangerous" = "Not so\ndangerous", "dangerous" = "Dangerous", "very dangerous" = "Very\ndangerous", "don't\nknow" = "Don't know")) +
  theme_minimal()

dist_PercRiskDUIC

#PERCRISK_DUIAmod_bin
data <- data %>% mutate(DUICDUIA.PERCRISK.2 = factor(DUICDUIA.PERCRISK.2, levels = c("ungefährlich", "nicht so gefährlich", "gefährlich", "sehr gefährlich", "weiß nicht"), labels = c("not dangerous", "not so dangerous", "dangerous", "very dangerous", "don't know")))

dist_PercRiskDUIA <- ggplot(data, aes(x = DUICDUIA.PERCRISK.2, fill = PERCRISK_DUIAmod_bin)) +
  geom_bar() +
  labs(x = "How dangerous do you think it is\nto drink a little more than is allowed before driving?", y = "Count", fill = "Dichotomized\nRisk\nPerception\nDUIA") +
  scale_fill_discrete(labels = c("high risk perc" = "High", "low risk perc" = "Low")) +
  scale_x_discrete(labels = c("not dangerous" = "Not\ndangerous", "not so dangerous" = "Not so\ndangerous", "dangerous" = "Dangerous", "very dangerous" = "Very\ndangerous", "don't\nknow" = "Don't know")) +
  theme_minimal()

dist_PercRiskDUIA

#BSSS
dist_BSSS <- ggplot(data, aes(x = BSSS_total, fill = BSSS_binary)) +
  geom_histogram(bins = 20) +
  #vertical line at median
  geom_vline(xintercept = median(data$BSSS_total), linetype = "dashed", color = "red", linewidth = 1) +
  labs(x = "Brief Sensation Seeking Scale - Score", y = "Count", fill = "Dichotomized\nSensation\nSeeking") +
  theme_minimal()

dist_BSSS

#speeding
dist_speed <- ggplot(data, aes(x = speeding_total, fill = speeding_binary)) +
  geom_histogram(bins = 20) +
  #vertical line at median
  geom_vline(xintercept = median_speeding_total, linetype = "dashed", color = "red", linewidth = 1) +
  labs(x = "Speeding Scale - Score", y = "Count", fill = "Dichotomized\nSpeeding") +
  theme_minimal()

dist_speed

#mental illness
dist_mentalillness <- ggplot(data, aes(x = DISTRESS_total, fill = mental_illness)) +
  geom_histogram(bins = 20) +
  labs(x = "Kessler Psychological Distress Scale - Score", y = "Count", fill = "Mental Illness") +
  theme_minimal()

dist_mentalillness
```


```{r save_plots, include=FALSE}
#save plots with graph2doc
graph2doc(x = dist_CAST, file = file.path(folder_plot, "dist_CAST.png"), width = 6, height = 4)
graph2doc(x = dist_PercRiskDUIC, file = file.path(folder_plot, "dist_PercRiskDUIC.png"), width = 6, height = 4)
graph2doc(x = dist_PercRiskDUIA, file = file.path(folder_plot, "dist_PercRiskDUIA.png"), width = 6, height = 4)
graph2doc(x = dist_BSSS, file = file.path(folder_plot, "dist_BSSS.png"), width = 6, height = 4)
graph2doc(x = dist_speed, file = file.path(folder_plot, "dist_speed.png"), width = 6, height = 4)
graph2doc(x = dist_mentalillness, file = file.path(folder_plot, "dist_mentalillness.png"), width = 6, height = 4)
```

# Correlations between variables (Cramers V)

```{r check correlations}

all_vars <- c("DUIA12m", "DUICINTOX12m", "CUD_CAST", "mental_illness", "medic_withprescr_only", "LEGALPROBL", "DUICPEERS12m", "BSSS_binary", "speeding_binary", "PERCRISK_DUICmod_bin", "PERCRISK_DUIAmod_bin", "urb_lab_en_bin", "edu_group", "sex", "agegroup")

# function to calculate Cramers V
catcorrm <- function(vars, dat) {
  if (!all(vars %in% colnames(dat))) stop("some variables are not in the data frame")
  n <- length(vars)
  cramer_matrix <- matrix(NA, nrow = n, ncol = n, dimnames = list(vars, vars))

  for (i in seq_along(vars)) {
    for (j in seq_along(vars)) {
      table_data <- table(dat[[vars[i]]], dat[[vars[j]]])
      cramer_matrix[i, j] <- assocstats(table_data)$cramer
    }
  }
  
  return(cramer_matrix)
}

# call function
cramers_v_matrix <- catcorrm(all_vars, as.data.frame(data))
cramers_v_df <- as.data.frame(as.table(cramers_v_matrix)) %>%
  rename(CramersV = Freq)

cramers_v_df
```


# Fit LCA Modell

```{r fit LCA model}

# model with covariates
f_with_cov <- with(data, cbind(DUICINTOX12m, DUIA12m, DUICPEERS12m, medic_withprescr_only, mental_illness, BSSS_binary, LEGALPROBL, CUD_CAST, PERCRISK_DUICmod_bin, PERCRISK_DUIAmod_bin, speeding_binary) ~ sex + agegroup + edu_group + urb_lab_en_bin)

# model without covariates
f <- with(data, cbind(DUICINTOX12m, DUIA12m, DUICPEERS12m, medic_withprescr_only, mental_illness, BSSS_binary, LEGALPROBL, CUD_CAST, PERCRISK_DUICmod_bin, PERCRISK_DUIAmod_bin, speeding_binary) ~ 1)



fit_lca_models <- function(data, formula, max_groups, nrep = 20, maxiter = 4000) {
  set.seed(12345)
  
  # Initialize a dataframe to store the results
  results <- data.frame(Model = integer(max_groups),
                        log_likelihood = numeric(max_groups),
                        resid_df = integer(max_groups),
                        BIC = numeric(max_groups),
                        AIC = numeric(max_groups),
                        likelihood_ratio = numeric(max_groups),
                        stringsAsFactors = FALSE)
  
  models <- list()
  min_bic <- 100000  # Initial value for the minimum BIC
  best_lca_model <- NULL
  
  # Loop to fit models with different numbers of classes
  for (i in 1:max_groups) {
    cat(sprintf("Fitting LCA model with %d classes...\n", i))
    
    tryCatch({
      # fit the LCA model
      lc <- poLCA(formula, data, nclass = i, na.rm = FALSE, nrep = nrep, maxiter = maxiter)
      
      # save the model fit statistics
      results[i, 1:6] <- c(i, lc$llik, lc$resid.df, lc$bic, lc$aic, lc$Gsq)
      
      # save the model in the list
      models[[i]] <- lc
      
      # update the best model based on the BIC value
      if (lc$bic < min_bic) {
        min_bic <- lc$bic
        best_lca_model <- lc
      }
    }, error = function(e) {
      # error handling in case model fitting fails
      print(paste("Error in model with", i, "classes"))
      results[i, ] <- c(i, NA, NA, NA, NA, NA)
    })
  }
  
  # return the results and the best model
  return(list(results = results, best_model = best_lca_model, models = models))
}

# fit models
lca_results <- fit_lca_models(data, f_with_cov, max_groups = 5)
lca_results_wo_cov <- fit_lca_models(data, f, max_groups = 5)

print(lca_results$results)
print(lca_results_wo_cov$results)
```

```{r optional: rerun best model to define class order}
# Reorder classes if needed
reorder_classes_if_needed <- function(lca_results, f_with_cov, data) {
  classprobs <- lca_results$best_model$P
  sorted_classprobs <- sort(classprobs, decreasing = TRUE)
  #check if classes are ordered correctly (i.e. in descending order of class size)
  if (all(classprobs == sorted_classprobs)) {
    print("Classes are ordered correctly")
    return(lca_results$best_model)
  } else {
    print("Classes are not ordered correctly")
    print("Reordering classes...")
    
    #rerun the best model in order to define the ordering of the classes depending on their size (the biggest class should be class 1,
    # so it defines the reference class in the latent class regression) 
    probs.start <- lca_results$best_model$probs.start
    lc_rerun <- poLCA(f_with_cov, data, nclass = 3, probs.start = probs.start, graphs = TRUE, na.rm = FALSE, maxiter = 4000)
    
    #reorder classes (INSERT desired order of classes)
    new.probs.start <- poLCA.reorder(probs.start, c(1, 3, 2))
    
    lc_rerun_2 <- poLCA(f_with_cov, data, nclass = 3, probs.start = new.probs.start, graphs = TRUE, na.rm = FALSE, maxiter = 4000)
    return(lc_rerun_2)
  }
}

lca_results_reordered <- reorder_classes_if_needed(lca_results, f_with_cov, data)
lca_results_wo_cov_reordered <- reorder_classes_if_needed(lca_results_wo_cov, f, data)
```

# Estimated class population shares
```{r class_shares}  
bestmodel <- lca_results_reordered

classprobs <- bestmodel$P #same as round(colMeans(bestmodel$posterior)*100,2)
class_se <- bestmodel$P.se

# Calculate confidence intervals for class population shares
ci_multiplier <- 1.96  # Z-value for 95% confidence interval
ci_lower <- classprobs - ci_multiplier * class_se
ci_upper <- classprobs + ci_multiplier * class_se

# Create a data frame to store class shares and confidence intervals
class_shares <- data.frame(
  class = seq_along(classprobs),  # Generates a sequence from 1 to number of classes
  share = classprobs,
  ci_lo = ci_lower,
  ci_up = ci_upper
)
```

# Predicted class membership (by model posterior probabilities)
```{r pred_class}
table(bestmodel$predclass)
prop.table(table(bestmodel$predclass))
```


# Profile plot of best fitting LCA model (with covariates)
```{r profile plot}
# constants
class_levels <- c("1", "2", "3")
shape_values <- c("1" = 16, "2" = 17, "3" = 18)
color_values <- c("1" = "#1f77b4", "2" = "#ff7f0e", "3" = "#2ca02c")
linetype_values <- c("1" = "solid", "2" = "longdash", "3" = "dotted")
x_levels <- c("DUIA12m", "DUICINTOX12m", "CUD_CAST", "medic_withprescr_only", "LEGALPROBL", "DUICPEERS12m", "PERCRISK_DUIAmod_bin", "PERCRISK_DUICmod_bin", "speeding_binary", "BSSS_binary", "mental_illness")
label_mapping <- c(
  "DUIA12m" = "DUIA", 
  "DUICINTOX12m" = "DUIC",
  "CUD_CAST" = "CUD\nrisk",
  "medic_withprescr_only" = "Medical\nuse",
  "LEGALPROBL" = "Legal\nproblems",      
  "DUICPEERS12m" = "DUIC\npeers",          
  "PERCRISK_DUIAmod_bin" = "Low risk\nperception\nDUIA",
  "PERCRISK_DUICmod_bin" = "Low risk\nperception\nDUIC",
  "speeding_binary" = "Speeding", 
  "BSSS_binary" = "Sensation\nSeeking",     
  "mental_illness" = "Mental\nillness"
)


# data preparation for plotting
lcaModelProbs <- reshape2::melt(bestmodel$probs, level = 1) %>%
  rename(class = Var1) %>%
  mutate(class = str_extract(class, "\\d+"),
         class = factor(class, levels = class_levels))

# add standard errors
probs_se_melted <- melt(bestmodel$probs.se, level = 1) %>%
  rename(value_se = value, class = Var1) %>%
  dplyr::select(-Var2) %>%
  mutate(value_se = round(value_se, digits = 10)) %>%
  distinct() %>%
  mutate(class = factor(class, levels = class_levels))

# add confidence intervals
lcaModelProbs <- left_join(lcaModelProbs, probs_se_melted, by = c("class", "L1")) %>%
  mutate(CI_lower = pmax(value - 1.96 * value_se, 0), # CI cannot be negative
         CI_upper = pmin(value + 1.96 * value_se, 1) # CI cannot be > 1
  )

# which variable manifestation probabilities to plot
lcaModelProbs_plotting <- lcaModelProbs %>% filter(Var2 %in% c("high", "yes", "low risk perc"))


# size of the classes in percent
class_shares <- sprintf("%.1f%%", bestmodel$P * 100)
label_values <- c(
  "1" = paste("Class 1\n", "(", class_shares[1], ")", "\n", sep = ""),
  "2" = paste("Class 2\n", "(", class_shares[2], ")", "\n", sep = ""),
  "3" = paste("Class 3\n", "(", class_shares[3], ")", "\n", sep = "")
)

# generate plot
LCA_profile_plot <- ggplot(lcaModelProbs_plotting, aes(x = L1, y = value, group = class, 
                                                       color = class, linetype = class)) +
    geom_point() +
    geom_line(linewidth=1) +
    geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.55, linewidth = 0.72, linetype = "solid") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(hjust = 0.5, vjust = 1),
      axis.text.y = element_text(size = 12),              
      axis.title.x = element_text(size = 14),          
      axis.title.y = element_text(size = 14),            
      plot.title = element_text(hjust = 0.5, size = 16), 
      legend.text = element_text(size = 12),             
      legend.title = element_text(size = 14),
      legend.position = "bottom",
      panel.grid.major.x = element_blank(),
      panel.grid.minor.y = element_blank()) +             
    scale_x_discrete(labels = label_mapping, limits = x_levels) +
    scale_color_manual(values = color_values, labels = label_values) +
    scale_linetype_manual(values = linetype_values, labels = label_values) +
    scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +
    labs(x = "", y = "Item Response Probability", color = "", linetype = "")

LCA_profile_plot

# save plot
ggsave(LCA_profile_plot, file = file.path(folder_plot, "LCA_profile_plot.png"), width = 11, height = 6, bg = "white")
graph2doc(x = LCA_profile_plot, file=file.path(folder_plot, "LCA_profile_plot.png"), width=10.5, height=6)
```

# Profile plot of best fitting LCA model (without covariates)
```{r profile plot without covariates}
# data preparation for plotting
bestmodel_wo_cov <- lca_results_wo_cov_reordered

lcaModelProbs_wo_cov <- reshape2::melt(bestmodel_wo_cov$probs, level = 1) %>%
  rename(class = Var1) %>%
  mutate(class = str_extract(class, "\\d+"),
         class = factor(class, levels = class_levels))

# add standard errors
probs_se_melted_wo_cov <- melt(bestmodel_wo_cov$probs.se, level = 1) %>%
  rename(value_se = value, class = Var1) %>%
  dplyr::select(-Var2) %>%
  mutate(value_se = round(value_se, digits = 10)) %>%
  distinct() %>%
  mutate(class = factor(class, levels = class_levels))

# add confidence intervals
lcaModelProbs_wo_cov <- left_join(lcaModelProbs_wo_cov, probs_se_melted_wo_cov, by = c("class", "L1")) %>%
  mutate(CI_lower = pmax(value - 1.96 * value_se, 0), # CI cannot be negative
         CI_upper = pmin(value + 1.96 * value_se, 1) # CI cannot be > 1
  )

#which variable manifestation probabilities to plot
lcaModelProbs_wo_cov_plotting <- lcaModelProbs_wo_cov %>% filter(Var2 %in% c("high", "yes", "low risk perc"))

# size of the classes in percent
class_shares <- sprintf("%.1f%%", bestmodel_wo_cov$P * 100)
label_values <- c(
  "1" = paste("Class 1\n", "(", class_shares[1], ")", "\n", sep = ""),
  "2" = paste("Class 2\n", "(", class_shares[2], ")", "\n", sep = ""),
  "3" = paste("Class 3\n", "(", class_shares[3], ")", "\n", sep = "")
)

#generate plot
LCA_profile_plot_wo_cov <- ggplot(lcaModelProbs_wo_cov_plotting, aes(x = L1, y = value, group = class, 
                                                                    color = class, linetype = class)) +
  geom_point() +
  geom_line(linewidth=1) +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), width = 0.2, alpha = 0.55, linewidth = 0.72, linetype = "solid") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(hjust = 0.5, vjust = 1, size = 12),
    axis.text.y = element_text(size = 12),              
    axis.title.x = element_text(size = 14),          
    axis.title.y = element_text(size = 14),            
    plot.title = element_text(hjust = 0.5, size = 16), 
    legend.text = element_text(size = 12),             
    legend.title = element_text(size = 14),
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.y = element_blank()) +             
  scale_x_discrete(labels = label_mapping, limits = x_levels) +
  scale_color_manual(values = color_values, labels = label_values) +
  scale_linetype_manual(values = linetype_values, labels = label_values) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +
  labs(x = "", y = "Item Response Probability", color = "", linetype = "")

LCA_profile_plot_wo_cov

# save plot
ggsave(LCA_profile_plot_wo_cov, file = file.path(folder_plot, "LCA_profile_plot_wo_cov.png"), width = 11, height = 6, bg = "white")
graph2doc(x = LCA_profile_plot_wo_cov, file=file.path(folder_plot, "LCA_profile_plot_wo_cov.png"), width=10.5, height=6)

```

# Model Fit

```{r check model fit}
lca_fitmetric_allmodels <- lca_results$results %>%
  gather(key = "Metric", value = "Value", -Model)

ggplot(lca_fitmetric_allmodels, aes(x = Model, y = Value, color = Metric)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(title = "Model Fit", x = "Number of Classes", y = "Value", color = "Metric") +
  theme_minimal()
```


# Diagnostic Criteria

## (Relative) Entropy

Relative entropy values above .8 indicate an acceptable classification of individuals into classes (Clark & Muthén, 2009). Values close to 1 are ideal.

A too low entropy value might indicate that the model is overly simplified and does not adequately account for the variation in the data. However, a too high entropy value might also suggest that the model is too complex and possibly exhibits overfitting.

```{r entropy_rel}
entropy.R2 <- function(fit) {
  entropy <- function(p) { # entropy for a probability vector
    sum(-p * log(p)) #minus sign ensures that the entropy is positive (as log(p) is negative) 
  }
  error_prior <- entropy(fit$P) # entropy of  estimated class proportions
  error_post <- mean(apply(fit$posterior, 1, entropy)) # calculate the entropy for each row (each observation), where each column represents a class. Then take the mean of these entropies.
  R2_entropy <- (error_prior - error_post) / error_prior
  cat("R2 Entropy: ", R2_entropy, "\n")
  cat("Error Prior: ", error_prior, "\n")
  cat("Error Posterior: ", error_post, "\n")
}

#entropy for best fitting model with covariates
entropy.R2(bestmodel)

#entropy for all models without covariates
entropy_results <- lapply(lca_results_wo_cov$models[1:5], entropy.R2)
```

## Average posterior class membership probabilities

- provide insights into how well the model assigns individuals to their most likely class
- are provided for their most likely assigned class, with values > .70 indicating well-separated classes (Nagin, 2005).

```{r posterior probabilities}

calculate_classification_metrics <- function(model, num_classes) {
  # Extract the posterior probabilities
  class_probs <- data.frame(model$posterior)

  # Calculate the maximum probability for each participant
  class_probs$max_prob <- apply(class_probs, 1, max)

  # Calculate average posterior probability
  avg_posterior_prob <- mean(class_probs$max_prob)
  cat("Average posterior probability: ", avg_posterior_prob, "\n")

  # Create a dataframe with posterior probabilities and predicted classes
  posteriors <- data.frame(model$posterior, model$predclass)

  # Rename the predclass column
  names(posteriors)[names(posteriors) == "model.predclass"] <- "predclass"

  # Calculate the classification table by summing the probabilities
  classification_table <- posteriors %>%
    group_by(predclass) %>%
    summarize(across(.cols = 1:num_classes, .fns = sum))

  # Transform the counts to proportions
  classification_table_prop <- classification_table %>%
    mutate(across(.cols = 2:(num_classes + 1), .fns = function(x) x/sum(x)))

  # Print the proportions table
  print(classification_table_prop)

  # Convert to matrix and calculate classification error
  classification_values_only <- as.matrix(classification_table[2:(num_classes + 1)])
  total_classification_error <- 1 - sum(diag(classification_values_only)) / sum(classification_values_only)

  cat("Total classification error: ", total_classification_error, "\n")

  # Return results as a list for potential further use
  return(list(avg_posterior_prob = avg_posterior_prob,
              classification_table = classification_table,
              classification_table_prop = classification_table_prop,
              total_classification_error = total_classification_error))
}

classification_metrics_wo_cov <- lapply(1:5, function(i) {
  #print number of subjects in each class
  print(table(lca_results_wo_cov$models[[i]]$predclass))
  calculate_classification_metrics(lca_results_wo_cov$models[[i]], num_classes = i)
})


calculate_classification_metrics(bestmodel, 3)

```

# Does the covariate model have a significant improvement in likelihood over the no-covariate model? 

```{r cov vs no cov}
#values of model without covariates
ll_wo_cov <- bestmodel_wo_cov$llik #log-likelihood
param_wo_cov <- bestmodel_wo_cov$npar # number of parameters

#values of model with covariates
ll_with_cov <- bestmodel$llik #log-likelihood
param_with_cov <- bestmodel$npar # number of parameters

#perfrom likelihood ratio test
G2 <- -2 * (ll_wo_cov - ll_with_cov) # G2 statistic (likelihood ratio test statistic)
df <- param_with_cov - param_wo_cov  # degrees of freedom (i.e. number of additional parameters in the model with covariates)
p_value <- pchisq(G2, df, lower.tail = FALSE)

cat("Likelihood ratio test results:\n")
cat("G2: ", G2, "\n")
cat("Degrees of freedom: ", df, "\n")
cat("p-value: ", p_value, "\n")

# yes, model with covariates has a significant improvement in likelihood over the no-covariate model
```

# Latent Class Regression Coefficients

```{r regression coefficients}
coefficients <- bestmodel$coeff  # log odds
se <- bestmodel$coeff.se         # standard errors

odds_ratios <- exp(coefficients)

#95% confidence intervals
lower_ci <- exp(coefficients - 1.96 * se) 
upper_ci <- exp(coefficients + 1.96 * se) 

odds_ratios
lower_ci
upper_ci

covariate_names <- rownames(coefficients)

# generate data frame for plotting
plot_data <- data.frame(
  Covariate = rep(rownames(coefficients), times = ncol(coefficients)),
  Class = rep(c("'DUIC-specific risk'-class (N = 151)", "'Global risk'-class (N = 129)"), each = nrow(coefficients)),
  OddsRatio = as.vector(odds_ratios),
  LowerCI = as.vector(lower_ci),
  UpperCI = as.vector(upper_ci)
)

plot_data <- plot_data %>% filter(Covariate != "(Intercept)")

# extra dataframe for headings
grouping_rows <- data.frame(
  Covariate = c("Gender (Reference: male)", "Agegroup (Reference: 18-24)", 
                "Education (Reference: low)", "Urbanicity (Reference: rural)"),
  Class = rep(c("'DUIC-specific risk'-class (N = 151)", "'Global risk'-class (N = 129)"), each = 2),
  OddsRatio = rep(NA, 4), 
  LowerCI = rep(NA, 4),  
  UpperCI = rep(NA, 4)
)

# bind rows with headings to the data frame
plot_data <- rbind(grouping_rows, plot_data)

plot_data$Covariate <- factor(plot_data$Covariate,
                              levels = c("sexweiblich", "Gender (Reference: male)", 
                                         "agegroup45-64", "agegroup35-44", "agegroup25-34", 
                                          "Agegroup (Reference: 18-24)", # heading for age groups
                                         "edu_grouphigh", "edu_groupmid", 
                                          "Education (Reference: low)",  # heading for education groups
                                         "urb_lab_en_binurban", "Urbanicity (Reference: rural)"))  # heading for urbanicity groups

# change label names
plot_data$Covariate <- factor(plot_data$Covariate,
                              labels = c("female", "Gender (Reference: male)", 
                                 "45-64", "35-44", "25-34", "Agegroup (Reference: 18-24)",
                                 "high", "medium", "Education (Reference: low)",
                                 "urban", "Urbanicity (Reference: rural)")) 
                              
LCARegCoeff <- ggplot(plot_data, aes(x = Covariate, y = OddsRatio)) +
  geom_point(aes(color = ifelse(OddsRatio > 1, "Above 1", "Below 1")), size = 2.5) +
  geom_errorbar(aes(ymin = LowerCI, ymax = UpperCI, color = ifelse(OddsRatio > 1, "Above 1", "Below 1")), width = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip(ylim = c(0.1, 6)) + 
  facet_wrap(~ Class) +
  scale_y_continuous(breaks = seq(0, 6, by = 1), labels = seq(0, 6, by = 1)) +
  geom_label(data = plot_data %>% filter(LowerCI > 1 | UpperCI < 1), 
            aes(label = paste0("OR = ", round(OddsRatio, 2)), vjust = -0.5, hjust = -0.2), 
            color = "black", size = 5, label.size = 0.5, fill = "white", label.padding = unit(0.15, "lines")) +
  theme_minimal(base_size = 20) +
  labs(title = "Odds of Membership in a DUIC-risk-class", subtitle = "Reference: 'Low risk'-class (N = 267), total N = 547", x = "", y = "Odds Ratio") + 
  scale_color_manual(values = c("Above 1" = "darkred", "Below 1" = "skyblue")) +
  theme(legend.position = "none", 
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.text = element_text(face = "bold"),
        panel.spacing = unit(0.8, "lines"),
        plot.margin = margin(0.5, 0.5, 0.5, 0.5, "cm"),
        axis.text.y = element_text(face = ifelse(levels(plot_data$Covariate) %in% c(
          "Gender (Reference: male)", "Agegroup (Reference: 18-24)", 
          "Education (Reference: low)", "Urbanicity (Reference: rural)"), "bold", "plain")))

# save plot
graph2doc(x = LCARegCoeff, file=file.path(folder_plot, "LCARegCoeff.png"), width=12, height=7)
```